---
title: "Amplicon tutorial"
author: "Yong-Xin Liu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Introduction

## Installation

The [amplicon R package](https://github.com/microbiota)
facilitates statistics and visualization of amplicon profiling data, in
particular 16S and ITS alpha / beta taxonomy, taxonomic profiling.

```{r install, eval=FALSE, include=TRUE}
# library(BiocManager)
# if (!requireNamespace("BiocManager", quietly=TRUE)){
#     install.packages("BiocManager")}
# if (!requireNamespace("BiocManager", quietly=TRUE)){
#     BiocManager::install("amplicon")}
```

Then load the package in R

```{r loading}
library(amplicon)  
```


## Data

The amplicon package is similar with
[phyloseq](https://github.com/joey711/phyloseq) data format. Mainly 
includes an OTU table (feature abundances), sample metadata (group, date,
site, ...), taxonomy table (annotation of OTUs in 7 levels taxonomy: 
Kingdom, Phylum, Class, Order, Family, Genus, Species).

Showing the example data format

**OTU table**

A table include reads counts of each OTU in each sample. Row name is OTU ID, and Column name is sample ID.

```{r otutab}
data(otutab)
dim(otutab)
otutab[1:3, 1:3]
```


**metadata**

A table include sample information. Row name must be sample ID, and as following BarcodeSequence, group ID, date, location / site, et. al. .

```{r metadata}
data(metadata)
head(metadata, n = 3)
```

**taxonomy**

A table include 7 levels of taxonomy of each OTU.

```{r taxonomy}
data(taxonomy)
head(taxonomy, n = 3)
```

# Diversity

Setting the basic parameter

```{r parameter, warning=FALSE}
# colnames of group ID in metadata
# 设置实验分组列名
group = "Genotype"
# Output figure width and height  
# Letter纸图片尺寸为单栏89 mm，双栏183 mm，页面最宽为247 mm
# 推荐比例16：10，即半版89 mm x 56 mm; 183 mm x 114 mm
width = 89
height = 59
```


## Alpha diversity boxplot

Plotting each figure mainly include three step:

1. Reading data and viewing format;
2. Plotting and visualization;
3. Saving figure.

```{r alpha_boxplot, fig.show='asis', fig.width=4, fig.height=2.5, warning=FALSE}
# Data reading
metadata = read.table("metadata.txt", header=T, row.names=1, sep="\t", comment.char="")
head(metadata, n = 3)
alpha_div = read.table("alpha_div.txt", header=T, row.names=1, sep="\t", comment.char="")
head(alpha_div, n = 3)
# capitalize
library(Hmisc)
colnames(alpha_div) = capitalize(colnames(alpha_div))
colnames(alpha_div)

# Plotting alpha diversity richness boxPlotting and statistics
(p = alpha_boxplot(alpha_div, index = colnames(alpha_div)[9], metadata, groupID = group))

# Saving figure
# 保存图片，大家可以修改图片名称和位置，长宽单位为毫米
ggsave("alpha_boxplot.pdf", p, width = width, height = height, units = "mm")
p1 = p

# 尝试探索不同的多样性各类
colnames(alpha_div)
alpha_boxplot(alpha_div, index = colnames(alpha_div)[12], metadata, groupID = group)

# 尝试不同的分组方式
colnames(metadata)
alpha_boxplot(alpha_div, index = colnames(alpha_div)[9], metadata, groupID = "Site")
```

## Alpha rarefraction curve 

```{r alpha_rare, fig.show='asis', fig.width=4, fig.height=2.5}
# Data reading
alpha_rare = read.table("alpha_rare.txt", header=T, row.names=1, sep="\t", comment.char="")
alpha_rare[1:3, 1:3]

# Plotting alpha rarefraction curve in group mean with standard error
(p = alpha_rare_curve(alpha_rare, metadata, groupID = group))

# Saving figure
ggsave("alpha_rarefraction_curve.pdf", p, width = width, height = height, units = "mm")
p2 = p
```

## Beta diversity —— principal coordinat analysis (PCoA)

```{r beta_pcoa, fig.show='asis', fig.width=4, fig.height=2.5}
# Data reading
beta_bray_curtis = read.table("beta_bray_curtis.txt", header=T, row.names=1, sep="\t", comment.char="")
beta_bray_curtis[1:3, 1:3]

# Plotting alpha rarefraction curve in group mean with standard error
(p = beta_pcoa(beta_bray_curtis, metadata, groupID = group))

# Saving figure
ggsave("beta_pcoa_bray.pdf", p, width = width, height = height, units = "mm")
p3 = p
```

## Beta diversity —— Constrained PCoA

```{r beta_cpcoa, fig.show='asis', fig.width=4, fig.height=2.5}
# Data reading
otutab = read.table("otutab.txt", header=T, row.names=1, sep="\t", comment.char="")
otutab[1:3, 1:3]

# Plotting alpha rarefraction curve in group mean with standard error
(p = beta_cpcoa(otutab, metadata, dis="bray", groupID = group))

# Saving figure
ggsave("beta_cpcoa_bray.pdf", p, width = width, height = height, units = "mm")
p4 = p
```

## Combo plots to published-ready figure

```{r div_combo, fig.show='asis', fig.width=6, fig.height=4}
library(cowplot)
(p0 = plot_grid(p1, p2, p3, p4, labels = c("A", "B", "C", "D")))
ggsave("diversity.pdf", p0, width = width * 2, height = height * 2, units = "mm")
```

